---
title: A Caribbean coastal lagoon serves as spawning habitat for multiple Callinectes spp.
output:
  pdf_document: default
  word_document: default
date: "2025-12-12"
---

---
title: "A Caribbean coastal lagoon serves as spawning habitat for multiple Callinectes spp."
format: pdf
editor: visual
author: "Olivia Pares"
---

```         
```

#Setting up directory

```{r}
 setwd("~/Desktop/Repository for manuscript 1/data")
```

# Libraries

```{r}
library(readxl)
library(tidyverse)
library(lubridate) 
library(car)
library(emmeans)
library(mgcv)
library(gamlss)
library(dplyr)
library(MASS)
library(ggplot2)
library(gamlss.tr)
library(here)
```

# Data Import

```{r}
## Load all committee CSVs back into R (expects the folder you created)
in_dir <- "data"

d1                         <- read.csv(file.path(in_dir, "d1.csv"),                         stringsAsFactors = TRUE,check.names = FALSE)
crabdata                   <- read.csv(file.path(in_dir, "crabdata.csv"),                   stringsAsFactors = T,check.names = FALSE)
sexsap                     <- read.csv(file.path(in_dir, "sexsap.csv"),                     stringsAsFactors = T,check.names = FALSE)
hatch                      <- read.csv(file.path(in_dir, "hatch.csv"),                      stringsAsFactors = T,check.names = FALSE)
sizedf <- read.csv(file.path(in_dir, "sizedf.csv"),
                   stringsAsFactors = FALSE,
                   check.names = FALSE)

## (Optional) verify they loaded
sapply(list(
  d1=d1, crabdata=crabdata, sexsap=sexsap, hatch=hatch, sizedf=sizedf
), nrow)


```

# Temporal and Spatial Catch

The following models evaluate species-specific and pooled Callinectes catch across stations and quarterly sampling periods.

```{r}

d1$sq <- as.factor(d1$sq)
d1$Year_Month <- as.factor(d1$Year_Month)
d1$Station <- as.factor(d1$Station)
d1$lagoon <- as.factor(d1$lagoon)
d1$Date <- as.Date(d1$Date) 

#Callinectes sapidus 
M1 <- gamlss(cpue.sap~re(fixed = ~ Year_Month + Station,
                           random = ~1|lagoon,
                           correlation = corAR1(form = ~ Date |lagoon/sq)
                           ,method = "REML")
             ,family=NBI()
             ,data=d1)

summary(getSmo(M1))
plot(M1)
anova(getSmo(M1))



```

```{r}
#Callinectes bocourti 
M2 <- gamlss(cpue.boc~re(fixed = ~ Year_Month + Station,
                           random = ~1|lagoon,
                           correlation = corAR1(form = ~ Date |lagoon/sq)
                           ,method = "REML")
             ,family=NBI()
             ,data=d1)

summary(getSmo(M2))
plot(M2)
anova(getSmo(M2))


```

```{r}

#Callinectes bocourti
M3 <- gamlss(cpue.boc~re(fixed = ~ Year_Month + Station,
                           random = ~1|lagoon,
                           correlation = corAR1(form = ~ Date |lagoon/sq)
                           ,method = "REML")
             ,family=NBI()
             ,data=d1)

summary(getSmo(M3))
plot(M3)
anova(getSmo(M3))


```

```{r}
#Callinectes similis 
M4 <- gamlss(cpue.sim~re(fixed = ~  Year_Month + Station ,
                           random = ~1|lagoon,
                           correlation = corAR1(form = ~ Date |lagoon/sq)
                           ,method = "REML")
             ,family=NBI()
             ,data=d1)

summary(getSmo(M4))
plot(M4)
anova(getSmo(M4))
```

```{r}
#Callinectes ornatus  
M5 <- gamlss(cpue.orn~re(fixed = ~  Year_Month + Station,
                           random = ~1|lagoon,
                           correlation = corAR1(form = ~ Date |lagoon/sq)
                           ,method = "REML")
             ,family=NBI()
             ,data=d1)

summary(getSmo(M5))
plot(M5)
anova(getSmo(M5))
```

```{r}
#Calculating total catch per sampling event
 d1<- d1 %>%
  group_by(Date, Station) %>%
  mutate(
    total_catch = cpue.sap + cpue.boc + cpue.sim + cpue.orn + cpue.lar + cpue.exa
  ) %>%
  ungroup()

#All Callinectes spp. catch modeled 
M5 <- gamlss(total_catch~re(fixed = ~  Station + Year_Month,
                           random = ~1|lagoon,
                           correlation = corAR1(form = ~ Date |lagoon/sq)
                           ,method = "REML")
             ,family=NBI()
             ,data=d1)

summary(getSmo(M5))
plot(M5)
anova(getSmo(M5))
```

## Size at Sexual Maturity

```         
Morphometric maturity was estimated for female crabs using abdomen width and carapace width relationships.
```

```{r}
#Package and libraries needed
#install.packages("sizeMat")
library(sizeMat)
```

#Determining the size at sexual maturity for *C. sapidus*

```{r}
sap<- (crabdata %>% filter(ID=="SAPIDUS"))
sap <- (sap%>% filter (sex=="F"))

#Factor for determining maturity

sap$`Abdomen width (mm)`
sap$`Abdomen width (mm)`<- as.numeric(sap$`Abdomen width (mm)`)

sap$carapace_width_spines <- as.numeric(sap$carapace_width_spines)

sap$abdomen_width <-sap$`Abdomen width (mm)`
sap$CW <- sap$carapace_width_spines

```

```{r}
#Classify only the female crabs
classify_sap_fem <- classify_mature(sap, varNames =c("CW", "abdomen_width"), 
varSex= "sex", selectSex= "F", method = "ld")

#Frequentist regression#
my_ogive_fq = morph_mature(classify_sap_fem, method = "fq", niter= 1000)
print(my_ogive_fq)

```

Plot the results of the size at sexual maturity of *C. sapidus* is determined to be 108.7 mm

```{r}

plot(my_ogive_fq, xlab = "Carapace width (mm)", ylab = "Proportion mature", col = c("blue", "red"),onlyOgive=TRUE)

legend("bottomright", legend = "C. sapidus", bty="n")



```

#Determining the size at sexual maturity for *C. bocourti*

```{r}
#Other crabs###


crabdata$carapace_width_spines <- as.numeric(crabdata$carapace_width_spines)
crabdata$`Abdomen width (mm)` <- as.numeric(crabdata$`Abdomen width (mm)`)
crabdata$abdomen_width <- crabdata$`Abdomen width (mm)`


boc<- (crabdata %>% filter(ID=="BOCOURTI"))
```

```{r}

classify_boc_fem <- classify_mature(boc, varNames =c("carapace_width_spines", "abdomen_width"), 
                                    varSex= "sex", selectSex= "F", method = "ld")

#Frequentist regression#

my_ogive_fq_boc = morph_mature(classify_boc_fem, method = "fq", niter= 1000)
print(my_ogive_fq)


plot(my_ogive_fq_boc, xlab = "Carapace width (mm)", ylab = "Proportion mature", col = c("blue", "red"),onlyOgive=TRUE)
legend("bottomright", legend = "C. bocourti", bty="n")


```

#Determining the size at sexual maturity for *C. ornatus*

```{r}
##Ornatus####
orn<- (crabdata %>% filter(ID=="ORNATUS"))

#Classify 
classify_orn_fem <- classify_mature(orn, varNames =c("carapace_width_spines", "abdomen_width"), 
                                    varSex= "sex", selectSex= "F", method = "ld")

#Frequentist regression#
my_ogive_fq_orn = morph_mature(classify_orn_fem, method = "fq", niter= 1000)
print(my_ogive_fq_orn)

#Plot for C. ornatus 
plot(my_ogive_fq_orn, xlab = "Carapace width (mm)", ylab = "Proportion mature", col = c("blue", "red"),onlyOgive=TRUE)


```

#Determining the size at sexual maturity for C. similis

```{r}

sim<- (crabdata %>% filter(ID=="SIMILIS"))

#Classify 
classify_sim_fem <- classify_mature(sim, varNames =c("carapace_width_spines", "abdomen_width"), 
                                    varSex= "sex", selectSex= "F", method = "ld")

#Frequentist regression#
my_ogive_fq_sim = morph_mature(classify_sim_fem, method = "fq", niter= 1000)
print(my_ogive_fq_sim)


plot(my_ogive_fq_sim, xlab = "Carapace width (mm)", ylab = "Proportion mature", col = c("blue", "red"),onlyOgive=TRUE)
legend("bottomright", legend = "C. similis", bty="n")


```

Too few data points to include *C. similis* and other low count species.

## Sex Ratio

The sex ratio was calculated for each sampling event and species as the number of males divided by the total number of males and females. This was then assessed to see if there was temporal or spatial variation in sex ratio across the different species.

# *Callinectes sapidus* sex ratio by Station

```{r}
#Sex ratio of Callinectes sapidus 

#Subset C. sapidus 
sexsap4 <- sexsap %>%
  filter(ID %in% c("C. similis"))

# Sum males and females for each station
sex_summary <- sexsap4 %>%
  group_by(Station) %>%
  summarise(
    Males = sum(Males),
    Females = sum(Females)
  )

sex_summary 
```

Analysis of sex ratio across stations

```{r}
# Convert to matrix for chi-square test
sex_table <- as.matrix(sex_summary[, c("Males", "Females")])

rownames(sex_table) <- sex_summary$Station

# View contingency table
sex_table

# Chi-square test of independence
chisq.test(sex_table)


# Fisher's exact test of independence when you have two nominal variables and you want to see whether the proportions of one variable are different depending on the value of the other variable. Use it when the sample size is small.
fisher.test(sex_table)
```

Pairwise comparisons to see how it differs by station

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(broom)

# total counts by station
station_counts <- sexsap4 %>%
  group_by(Station) %>%
  summarise(
    Males = sum(Males),
    Females = sum(Females),
    .groups = "drop"
  )

# Step 2:all pairwise combinations of stations
station_pairs <- combn(unique(station_counts$Station), 2, simplify = FALSE)

# Step 3: Run Fisher's Exact Test for each pair
pairwise_results <- map_dfr(station_pairs, function(pair) {
  subset_data <- station_counts %>%
    filter(Station %in% pair)
  
  # Create 2x2 matrix for Fisher's test
  test_matrix <- as.matrix(subset_data[, c("Males", "Females")])
  rownames(test_matrix) <- subset_data$Station
  
  # Run Fisher's test
  test <- fisher.test(test_matrix)
  
  tibble(
    Station1 = pair[1],
    Station2 = pair[2],
    p_value = test$p.value
  )
})

# Step 4: Adjust p-values for multiple testing
pairwise_results <- pairwise_results %>%
  mutate(p_adj_bonf = p.adjust(p_value, method = "bonferroni"),
         p_adj_fdr = p.adjust(p_value, method = "fdr"))

pairwise_results
```

\##*Callinectes sapidus* sex ratio by Quarter (Year_Month)

```{r}
# Sum males and females for each quarter
sex_summary <- sexsap4 %>%
  group_by(Year_Month) %>%
  summarise(
    Males = sum(Males),
    Females = sum(Females)
  )

sex_summary 
```

```{r}
# Convert to matrix for chi-square test
sex_table <- as.matrix(sex_summary[, c("Males", "Females")])

rownames(sex_table) <- sex_summary$Year_Month

# View contingency table
sex_table

# Chi-square test of independence
chisq.test(sex_table)


# Fisher's exact test of independence when you have two nominal variables and you want to see whether the proportions of one variable are different depending on the value of the other variable. Use it when the sample size is small.
fisher.test(sex_table)
```

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(broom)

# (Optional) sanity check
if (dplyr::n_distinct(sexsap4$Year_Month) != 5) {
  warning("Year_Month does not have exactly 5 levels in `sexsap4`.")
}

# Step 1: Create total counts by Year_Month (across all stations)
ym_counts <- sexsap %>%
  group_by(Year_Month) %>%
  summarise(
    Males   = sum(Males,   na.rm = TRUE),
    Females = sum(Females, na.rm = TRUE),
    .groups = "drop"
  )

# Step 2: Generate all pairwise combinations of Year_Month levels
ym_pairs <- combn(unique(as.character(ym_counts$Year_Month)), 2, simplify = FALSE)

# Step 3: Run Fisher's Exact Test for each Year_Month pair
pairwise_results <- map_dfr(ym_pairs, function(pair) {
  pair_chr <- as.character(pair)

  subset_data <- ym_counts %>%
    filter(Year_Month %in% pair_chr) %>%
    mutate(Year_Month = as.character(Year_Month)) %>%
    arrange(factor(Year_Month, levels = pair_chr))

  # Create 2x2 matrix (rows = the two Year_Month levels; cols = Males/Females)
  test_matrix <- as.matrix(subset_data[, c("Males", "Females")])
  rownames(test_matrix) <- subset_data$Year_Month

  test <- fisher.test(test_matrix)

  tibble(
    Year_Month1 = pair_chr[1],
    Year_Month2 = pair_chr[2],
    p_value     = test$p.value
  )
})

# Step 4: Adjust p-values for multiple testing
pairwise_results <- pairwise_results %>%
  mutate(
    p_adj_bonf = p.adjust(p_value, method = "bonferroni"),
    p_adj_fdr  = p.adjust(p_value, method = "fdr")
  )

pairwise_results

```

## Spawning Maturation Cycle

Estimate recovery time for an ovary between successive sponge production #Regress the average ovarian condition of ovigerous female per sponge color against the respective embryo age in days (d, where d = 0 at spawning) assuming that each successive sponge color represented 3 d of embryo development. Using the resulting regression equation, an average ovarian condition for inter-brood females was predicted This would be to determine if in the tropics female *C. sapidus* can produce more than 7 broods as seen in the **Ovarian development**: light , medium, full (1, 2, 3)

Sponge color:

Day 0-3 = Orange

Day 3 -6 = Dark Orange

Day 6 - 9 = Brown

Day 9 - 12 = Black

```{r}
H1<- hatch %>% 
  group_by(Embyro_day, Year_month,Station,ovigerous)%>%
  summarize(
    mean_ova_dev = mean(Ova_dev,na.rm=T),
    .groups = "drop"
  )

H1$Year_month <- as.factor(H1$Year_month)
H1$Station <- as.factor(H1$Station)

```

The Full Model

```{r}
mh1<- lm(mean_ova_dev ~ Embyro_day + Year_month + Station , data=H1)
par(mfrow = c(2, 2))

plot(mh1)

# Normality of residuals
shapiro.test(residuals(mh1))

# Homoscedasticity
library(car)
ncvTest(mh1)

summary(mh1)
```

Simplified model

```{r}
# Final models compared
step_model <- lm(mean_ova_dev ~ Embyro_day + Year_month, data = H1)
mh1        <- lm(mean_ova_dev ~ Embyro_day, data = H1)

# AICs and ΔAIC
AIC(step_model); AIC(mh1)
AIC(mh1) - AIC(step_model)

# Akaike weights
dAIC <- c(0, AIC(mh1) - AIC(step_model))           # step_model first
w    <- exp(-0.5 * dAIC) / sum(exp(-0.5 * dAIC))
w    # ~ c(0.508, 0.492)

# Nested-model F test (does Year_month help beyond Embyro_day?)
anova(mh1, step_model, test = "F")

# BIC check (parsimony criterion)
BIC(mh1, step_model)
```

Plot the model

```{r}
library(ggplot2)

ggplot(H1, aes(x = Embyro_day, y = mean_ova_dev, color = ovigerous, fill = ovigerous)) +
  geom_point(size = 3, shape = 21) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE) +
  labs(x = "Time (days)", y = "Average Ovarian Development") 
  
  # Style: clean, no grid, box outline
  #theme_bw(base_size = 14) +       # white background with black border
  #theme(
  #  panel.grid = element_blank(),  # removes all grid lines
  #  panel.border = element_rect(color = "black", fill = NA, size = 0.8), # adds box border
  #  legend.title = element_blank(),
   # legend.position = c(0.8, 0.2),
   # legend.background = element_rect(fill = "white", color = "black"),
   # axis.title = element_text(size = 14),
   # axis.text = element_text(size = 12)
  #) +
#  scale_color_manual(values = c("non-ovigerous" = "gray40", "ovigerous" = "black")) +
#  scale_fill_manual(values = c("non-ovigerous" = "white", "ovigerous" = "black"))


```

# Size Distribution of *Callinectes* spp.

```{r}
# Make boxplot
sizedf$`Carapace width (lateral spines) (mm)`

ggplot(sizedf, aes(x = sex, y = `Carapace width (lateral spines) (mm)`, fill = sex)) +
  geom_boxplot() +
  facet_wrap(~ID) +
  labs( x = "Sex",
    y = "Carapace Width (mm)"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )


```

# Length-weight relationship

```{r}

head(sizedf)
library(dplyr)
#Filter out to include only the female non-ovigerous C. sapidus 
femsap_size<- sizedf %>%
  filter(ID=="SAPIDUS",sex == "F", 
         is.na(Ovigerous) | Ovigerous ==0,Maturity != "IM")



min(femsap_size$`weight (g)`, na.rm=TRUE)

femsap_size$logL <- log(femsap_size$`Carapace length (mm)`)
femsap_size$logW <- log(femsap_size$`weight (g)`)

hist(femsap_size$logW)
hist(femsap_size$logL)

femsap_size$Year_Month <- as.factor(femsap_size$Year_Month)

lm1<- lm(logW~logL,data=femsap_size)
library(FSA)
summary(lm1)


#Log(W) = -5.07 + 2.39 * log(L)

#W = 0.0063CL^2.39 
#Multiple R-squared:  0.7388
#Adjusted R-squared:  0.7345
#p-value: < 2.2e-16

plot(femsap_size$`weight (g)`,femsap_size$`Carapace length (mm)`)

symbs<- c(1,16)
plot(`weight (g)` ~`Carapace length (mm)`, data = femsap_size, ylab="Weight (g)",xlab="Total carapace length (mm)", cex=0.6)

# Scatterplot
plot(
  `weight (g)` ~ `Carapace length (mm)`,
  data = femsap_size,
  ylab = "Weight (g)",
  xlab = "Total carapace length (mm)",
  pch = 16,
  cex = 0.6
)

# Add fitted power curve
L_seq <- seq(
  min(femsap_size$`Carapace length (mm)`, na.rm = TRUE),
  max(femsap_size$`Carapace length (mm)`, na.rm = TRUE),
  length.out = 200
)

W_fit <- 0.0063 * L_seq^2.39

lines(L_seq, W_fit, lwd = 2)



```
